{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2023-11-01 paper/si.tex
wip on si

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$, with $N$ being even.
The dynamics unfolds on two time-scales. 
The short time scale describes the game dynamics. 
Here the $N$ individuals are randomly matched to form $N/2$ pairs to interact in a repeated prisoner's dilemma.  
Each round, individuals can choose whether to cooperate ($C$) or defect ($D$). 
In the most general setting, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{align}
	\bordermatrix{%
		& C & D \cr
		C &\ R &\ S\  \cr
		D &\ T &\ P\ \cr
	} . \label{eq:prisoner}
\end{align}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action. 
Then the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\mathbf{s}\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On this short time scale, the players' strategies are fixed, and players are consecutively matched to play many repeated games with randomly changing interaction partners.

@@

@@

@@

\noindent
{\bf Impact of mutations.} 
All of our previous simulations are run under the assumption that mutations are rare,~$\mu\!\rightarrow\!0$. 
This allows us to use the formula~\eqref{eq:fixation_probability} for a mutant's fixation probability. 
While this limit allows for more efficient computation procedures, it rules out the possibility that two or three strategies might stably coexist in a population~\cite{tkadlec:pnas:2023}.
To explore the effect of positive mutation rates, we performed five independent runs of the pairwise process described in
Section~\ref{section:model} for different values of $\mu$.
At each time step we record the average cooperation of the resident population. 
The results for different mutation rates are shown in Figure~\ref{fig:mutation}. 
The cooperation rate is higher when players have perfect memory (compared to limited memory), regardless of the
mutation rate. 
Only as the mutation rate approaches one, and the processes become fully random, the cooperation rate approaches $1/2$ in both simulations.
>>>>>>> origin/dev

@@

@@

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography}
}

>>>>

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$, with $N$ being even.
The dynamics unfolds on two time-scales. 
The short time scale describes the game dynamics. 
Here the $N$ individuals are randomly matched to form $N/2$ pairs to interact in a repeated prisoner's dilemma.  
Each round, individuals can choose whether to cooperate ($C$) or defect ($D$). 
In the most general setting, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{equation}\label{eq:prisoner}
    \begin{blockarray}{ccc}
        & \text{cooperate} & \text{defect} \\
        \begin{block}{c(cc)}
            \text{cooperate} & R & S \\
            \text{defect} & T & P \\
        \end{block}
    \end{blockarray}.
\end{equation}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action. 
Then the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\mathbf{s}\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On this short time scale, the players' strategies are fixed, and players are consecutively matched to play many repeated games with randomly changing interaction partners.

@@

\begin{equation}\label{eq:expected_payoff}

@@

\subsection{Updates based on the realized payoff one repeated game}\label{section:one_interaction_full_memory}

{\bf Motivation.}
In the case of limited memory, we have thus far considered that individuals
update their strategies based solely on the number of interactions and by
considering only the payoff they received in those interactions in the final
turns. \\

\noindent
A player's payoff in such an interaction is calculated based on the
probabilities. In comparison, in the perfect memory case, individuals consider
their expected payoff from the interaction. A natural question to ask is what
happens if individuals update based on the expected payoffs of the interaction.

\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
We consider the population setup once more, consisting of \(k\) mutants with the
strategy \(\strategy_\mutant=(y_\mutant,p_\mutant,q_\mutant)\), and \(N-k\)
residents with strategies \(\strategy_\resident =
(y_\resident,p_\resident,q_\mutant)\). In each step of the evolutionary process,
we select a learner and a role model. The learner or the role model can be
either a resident, and the other player is a mutant. As mentioned previously,
two major cases must be considered:

\begin{enumerate}
  \item The learner and the role model have had their most recent interaction
  with each other.
  \item The learner's last interaction was not with the role model. In the
  latter case, there are four different subcases based on whether the resident's
  last interaction partner was a mutant or a resident, and whether the mutant's
  last interaction partner was a mutant or a resident.
\end{enumerate}

\noindent
In each pairing, we assume that the players interact for a specific number of
rounds, and the payoff from their interaction is the average of the payoffs
described in Eq.~\eqref{eq:expected_payoff}. As a result, there are four
possible payoffs in our population setup: the payoffs an individual with a
resident strategy can receive when interacting with an individual with the same
strategy or with a mutant strategy, and similarly, the payoffs for the mutant
strategy. \\

\noindent
After these preparations, we can define the probability that the number of
mutants either increases or decreases by one.

\begin{equation*}
  \setlength{\arraycolsep}{1pt} 
  \begin{array}{llll}
    \lambda^+	 &=
  &\frac{1}{N\!-\!1}\cdot  & \rho\left(\pi (\strategy_\resident, \strategy_\mutant), \pi (\strategy_\mutant, \strategy_\resident)\right)\\[0.5cm]
  &+
  &\frac{N\!-\!2}{N\!-\!1} \cdot 
  &\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3}\cdot \rho(\pi (\strategy_\resident, \strategy_\mutant), \pi (\strategy_\mutant, \strategy_\mutant)) + 
   \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3}\cdot \rho(\pi (\strategy_\resident, \strategy_\mutant), \pi (\strategy_\mutant, \strategy_\resident)) \right.\\[0.5cm]
  &&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} \cdot \rho(\pi (\strategy_\resident, \strategy_\resident), \pi (\strategy_\mutant, \strategy_\mutant)) + 
   \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} \cdot \rho(\pi (\strategy_\resident, \strategy_\resident), \pi (\strategy_\mutant, \strategy_\resident))\right],
  \end{array}
\end{equation*}

\begin{equation*}
  \setlength{\arraycolsep}{1pt} 
  \begin{array}{llll}
    \lambda^+	 &=
  &\frac{1}{N\!-\!1}\cdot  & \rho\left(\pi (\strategy_\mutant, \strategy_\resident), \pi(\strategy_\resident, \strategy_\mutant)\right)\\[0.5cm]
  &+
  &\frac{N\!-\!2}{N\!-\!1} \cdot 
  &\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3}\cdot \rho(\pi(\strategy_\mutant, \strategy_\mutant), \pi(\strategy_\resident, \strategy_\mutant)) + 
   \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3}\cdot \rho(\pi(\strategy_\mutant, \strategy_\resident), \pi(\strategy_\resident, \strategy_\mutant)) \right.\\[0.5cm]
  &&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} \cdot \rho(\pi(\strategy_\mutant, \strategy_\mutant), \pi(\strategy_\resident, \strategy_\resident)) + 
   \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} \cdot \rho(\pi(\strategy_\mutant, \strategy_\resident), \pi(\strategy_\resident, \strategy_\resident),)\right].
  \end{array}
\end{equation*}

\noindent
{\bf Stochastic stability of cooperation.}
We once again calculate how easily a single \alld{} mutant can invade into a
resident population of \gtft{} players. When resident interacts with another
resident or with a mutant,
the expected payoffs based on Eq.~\eqref{eq:expected_payoff} of the interactions are,
\begin{equation*}
\begin{array}{ll}
  \pi (\strategy_\resident, \strategy_\resident) =  b - c,
  &\pi (\strategy_\resident, \strategy_\mutant) =  - c \left(\delta \left(q - 1\right) + 1\right).
\end{array}
\end{equation*}
The expected payoff of the mutant, are
\begin{align*}
\begin{array}{ll}
  \pi (\strategy_\mutant, \strategy_\resident) =  b \left(\delta \left(q - 1\right) + 1\right),
  &\pi (\strategy_\mutant, \strategy_\mutant) =  0.
\end{array}
\end{align*}
Based on these expected payoffs, we compute the ratio
\begin{equation}
  \frac{\lambda^{-}_1}{\lambda^{+}_1} =
  \frac{\frac{1}{N-1} + \frac{N-2}{N-1}}{\frac{1}{N-1} + \frac{N-2}{N-1}}
\end{equation}

\begin{equation}
  \frac{\lambda^{-}_1}{\lambda^{+}_1} =
  \frac{1 + e^{- \beta \left(b \left(\delta q - \delta\right) + c\right)}}{1 + e^{\beta \left(b \left(\delta q - \delta\right) + c\right)}}.
\end{equation}
For stochastic stability of cooperation we require this ratio to be larger than one, which is equivalent to
\begin{equation} \label{Eq:Condition_OneInteractionExpectedPayoff}
  q<1 - \frac{c}{\delta b}
\end{equation}
Interestingly, this conditions is the same as the condition for stochastic
stability in the case of perfect memory, see
Eq.~\eqref{Eq:Condition_PerfectMemory}.

@@

\noindent
{\bf Impact of mutations.} 
All of our previous simulations are run under the assumption that mutations are rare,~$\mu\!\rightarrow\!0$. 
This allows us to use the formula~\eqref{eq:fixation_probability} for a mutant's fixation probability. 
While this limit allows for more efficient computation procedures, it rules out the possibility that two or three strategies might stably coexist in a population~\cite{tkadlec:pnas:2023}.
To explore the effect of positive mutation rates, we performed five independent runs of the pairwise process described in
Section~\ref{section:model} for different values of $\mu$.
At each time step we record the average cooperation of the resident population. 
The results for different mutation rates are shown in Figure~\ref{fig:mutation}. 
The cooperation rate is higher when players have perfect memory (compared to limited memory), regardless of the
mutation rate. 
Only as the mutation rate approaches one, and the processes become fully random, the cooperation rate approaches $1/2$ in both simulations.

@@

\begin{figure}[!htbp]
  \centering 
  \includegraphics[width=\textwidth]{static/cooperation_rate_over_b_and_beta_memory_one.pdf}
  \caption{{\bf Evolution of direct reciprocity for different parameter values.}}
\end{figure}

@@

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography.bib}
}
